Visualizing Independence Using Extended Association Plots

### 2003-05-22 trunk/vcd/paper/dsc2003.Rnw
paper for dsc proceedings added (old file dsc2003.tex deleted).
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@260 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
draft for dsc2003 proceedings paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@261 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt]{article}
\usepackage{Sweave, graphicx}
\usepackage{Z}

@@

\SweaveOpts{engine=R,eps=FALSE}
%\VignetteIndexEntry{Visualizing Independence Using Extended Association and Mosaic Plots}
%\VignetteDepends{vcd,MASS,grid}
%\VignetteKeywords{mosaic plot, association plot, independence, HCL colors}
%\VignettePackage{vcd}

\setkeys{Gin}{width=0.8\textwidth}

@@

\title{Visualizing Independence Using Extended Association and Mosaic Plots}
\author{\hfill Achim Zeileis$^\dag$ \hfill David Meyer$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
 Technische Universit\"at Wien}\\
\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
}
\date{}
\maketitle

\begin{abstract}
Two visualization techniques for the independence problem in 2-way
contingency tables---assocation and mosaic plots---are extended in two directions:
\begin{enumerate}
\item The visualization is enhanced by displaying the significance of an appropriate
test for independence and by using improved color schemes.
\item The implementation in the \textsf{R} system is improved using a more modular
design and allowing for more flexible specification of plotting parameters.
\end{enumerate}
Furthermore, some enhancements for displays of various types of independence in
general multi-way tables are suggested.
\end{abstract}

@@

Statistical models built for the analysis of multivariate data
quickly become complex with increasing dimensionality.
One idea of visualization techniques is to
use the human visual system to detect structures in the data that
possibly are not obvious from solely numeric output (e.g., test
statistics).
The \textsf{R} package \texttt{vcd}---inspired by the Book ``Visualizing Categorical
Data'' \citep{vcd:Friendly:2000}---includes methods for the (mostly
graphical) exploration of categorical data, such as:
\begin{itemize}
 \item fitting and graphing of discrete distributions,
 \item plots and tests for independence and symmetry problems,
 \item visualization techniques for log-linear models.
\end{itemize}
Here, we focus on the visualization of independence,
in particular in 2-way tables.

\section{Tests for independence in 2-way tables} \label{sec:tests}

We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
convenience the number of observations is denoted $n = n_{++}$.

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables.
%% reference needed

But this is not the only plausible way of aggregation of the Pearson
residuals. There are many conceivable functionals $\lambda(\cdot)$ which
lead to reasonable test statistics $\lambda(\{r_{ij}\})$, the sum of squares
is just one of them. Another functional suitable for identifying
the cells which cause the dependence (if any) is the maximum of the absolute
values
\begin{equation} \label{eq:MaxAbs}
M \quad = \quad \max_{i, j} |r_{ij}|.
\end{equation}
Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[][ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells causing
the dependence can easily be identified.

Furthermore, the main reason for using the unconditional limiting distribution
for the $\chi^2$ statistic (\ref{eq:ChiSquare}) was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
performing permutation tests---either by simulation or by computation of the
permutation distribution---became more and more popular. In particular for
the independence hypothesis (\ref{eq:H0}), using a permutation test is very intuitive
due to the permutation invariance (given row and column sums) of this problem.
By employing this approach the permutation distribution of statistics of
type $\lambda(\{r_{ij}\})$ (also including the $\chi^2$ statistic) can be derived.

Other classical tests for the hypothesis of independence  not fitting in
the Pearson residual-based framework described above include Fisher's exact test,
asymptotic tests of the odds ratio (for $2\times2$ tables) and the Mantel-Haenzel
test.

\section{Visualization techniques for 2-way tables} \label{sec:visualization}

The two best known visualization techniques for independence in 2-way tables are
association plots and mosaic plots. Both are suitable to bring out departure of an
observed table $(n_{ij})$ from the expected table $(\hat n_{ij})$ in a graphical way.
The former focuses on the visualization of the Pearson residuals $r_{ij}$ (under
independence) while the latter primarily displays the observed frequencies
$n_{ij}$.

\subsection{Association plots}

@@

Figure \ref{fig:goals-assoc-classic} shows the association plot for the table of
home and away goals for all 306 games in the 1995/6 season of the German soccer league
Bundesliga \citep[the data is available in the package \texttt{vcd}]{vcd:Knorr-Held:1999}.
In particular, it can be seen that there are
fewer games ending 1--0 and more ending 1--1 than would be expected under independence.
But this does not yet imply the significance of some test statistic.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-classic,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(goals, rot = 0, gp = gp.binary)
@
\caption{Association plot for the Bundesliga data.}
\label{fig:goals-assoc-classic}
\end{center}
\end{figure}

\subsection{Mosaic plots}

Mosaic plots can be seen as an extension of grouped
bar charts where width and height of the bars show the relative
frequencies of the two variables: a mosaic plot simply
consists of a collection of tiles whose
sizes are proportional to the observed cell frequencies
(see Figure \ref{fig:goals-mosaic-classic}).

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-classic,fig=TRUE,echo=FALSE,height=6,width=7>>=
mosaicplot(t(goals), shade = TRUE, main = "")
@
\caption{Friendly mosaic plot for the Bundesliga data.}
\label{fig:goals-mosaic-classic}
\end{center}
\end{figure}

Sequential horizontal and vertical recursive splits are used to
visualize the frequencies of more than two variables, each new variable
conditional to the previously entered variables. A first extension by
\cite{vcd:Friendly:1994} uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under independence. This approach does not only work in 2-way
tables but also in log-linear models fitted to multi-way tables. In this
extension, positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}. In particular,
this shading does not provide a visualization of the maximum test (\ref{eq:MaxAbs})
as in general it is unknown to which significance level $\alpha$ the
values $c_\alpha = 2$ and $4$ correspond for the table being visualized.

Figure \ref{fig:goals-mosaic-classic} shows the Friendly mosaic plot for
the same Bundesliga data as in Figure \ref{fig:goals-assoc-classic}.
In this plot, the residuals
are visualized by the color shading of the cells. Again, the cells for
the 1--0 and 1--1 results are highlighted and furthermore, there seem to be
too many 6--3 results than would be plausible under independence. Whereas
the association plot in Figure \ref{fig:goals-assoc-classic} contains
no information about the result of a test for independence, the shading
in this plot conveys the impression that there might be
significant dependence (which would indicate that there
is not only {\em no} home field advantage in the German Bundesliga in 1995/6 but even
an advantage for the away team). However, it is unclear whether there
is evidence for this or not.

\section{Extensions} \label{sec:extensions}

In this section, we suggest some improvements and extensions which can be
applied to both visualization techniques introduced in Section
\ref{sec:visualization}. First, the visual enhancements
\begin{itemize}
\item combining visualization of and testing for independence
\item improved color schemes
\end{itemize}
are addressed. Second, the implementation enhancements
\begin{itemize}
\item more modular implementation using \texttt{grid}
\item more flexible plotting parameter specification
\end{itemize}
are described. None of these should be thought of as finished work
but rather as a collection of ideas which still need thorough
refinement and proper implementation.

\subsection{Visualization enhancements}

The extensions of \cite{vcd:Friendly:1994} to mosaic plots provide
substantial improvement of the original mosaic plots and enhanced them
from a plot for contingency tables to a visualization technique for
log-linear models (and therefore also for certain independence problems).
However, it has two major drawbacks: First,
the significance level for the hard-coded critical values 2 and 4 is
usually unknown. Second, the colors for (light) blue and red
used in the Michael Friendly's \textsf{SAS} implementation and also in the base implementation
in \textsf{R} are not device-independent and
do not provide homogeneous saturations over different colors
and copier proofness. We suggest some ways to overcome these drawbacks.

As pointed out in Section \ref{sec:tests}, the critical values for
the maximum statistic $M$ from (\ref{eq:MaxAbs}) can be derived
from the permutation distribution. Instead of the hard-coded
values 2 and 4, the particular critical values, e.g., at the levels
$\alpha = 0.1$ and $0.01$, for the table to be visualized could be
used. Hence, exactly those residuals causing the (potential)
dependency within the table are highlighted. At the moment, these critical values
are derived in \texttt{vcd} by simulation of the underlying distribution.

The implementation of Friendly mosaic plots in \textsf{SAS}
uses (by default) colors in the HLS (Hue--Luminance--Saturation) color space
with hues for blue and red, full saturation, and varying luminance
for lighter colors. The current \textsf{R} implementation is
based on colors in the HSV (Hue--Saturation--Value) space where
decreasing the saturation from 1 towards 0
%% ---given the hue for blue or red and full saturation---
generates very similar colors
as increasing the luminance for HSL colors from 0.5 towards 1.
As an alternative to those HLS and HSV colors, the device-independent
HCL (Hue--Chroma--Luminance) color space \citep{vcd:Ihaka:2003} could be
used which also provides homogeneous saturations over different colors and copier
proofness. The resulting association plot (using critical values for
$\alpha = 0.1$ and $0.01$) for the Bundesliga data can be seen in
Figure \ref{fig:goals-assoc-signif}. It indicates very clearly that
there is no evidence for rejecting the hypothesis of independence
in this table. Thus, there seems to be neither an advantage for
the home team nor for the away team.

\setkeys{Gin}{width=0.8\textwidth}
\begin{figure}[htbp]
\begin{center}
<<goals-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(goals, rot = 0)
@
\caption{Extended association plots for the Bundesliga data.}
\label{fig:goals-assoc-signif}
\end{center}
\end{figure}

To illustrate what a `postive' example looks like we give the association plot
for the well-known Hair-Eye-Color data set (avilable in base \textsf{R}) which cross-tabulates the hair and eye
color and sex of 592 statistics students: Figure \ref{fig:haireye-assoc-signif}
shows the association plot for the 328 female students. It is easy to see that
the hypothesis of independence is rejected at 1\% level by the maximum test (\ref{eq:MaxAbs})
due to higher frequencies of students with hair and eye color
black/brown and blond/blue and lower frequencies of
blond/brown and brown/blue than would be plausible under independence. Furthermore,
it can also be seen that the residual corresponding to the black/blue
cell is significant at the 10\% level.

\begin{figure}[htbp]
\begin{center}
<<haireye-assoc-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.assocplot(haireye)
@
\caption{Extended association plots for the Hair-Eye-Color data (female students).}
\label{fig:haireye-assoc-signif}
\end{center}
\end{figure}

We have seen that this approach performs very well when the test statistic
which should be visualized is the maximum statistic (\ref{eq:MaxAbs}). However,
it is difficult to extend this approach to general tests for independence,
in particular in the HCL space, where the ranges of chroma and luminance are
not independent. In contrast the HSV space has the advantage that all
three dimensions hue, saturation and value have the range $[0,1]$ and can
be varied independently. Therefore, a different approach than the one
described above could be to use the hue for the sign of the residuals as
before, the saturation for the absolute size of the residuals, and the
value as an indicator for the significance of some test statistic (only
using the full value when the overall test for independence rejects
the null hypothesis). The resulting mosaic plots under this paradigm
are depicted in Figure \ref{fig:both-mosaic-signif}. Again, it can
clearly be seen that despite some large residuals there is no
evidence against independence for the Bundesliga data, but that the
null hypothesis has to be rejected for the Hair-Eye-Color data.

\begin{figure}[htbp]
\begin{center}
<<goals-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.mosaicplot(t(goals), main = NULL)
@
<<haireye-mosaic-signif,fig=TRUE,echo=FALSE,height=6,width=6>>=
grid.mosaicplot(t(haireye), main = NULL)
@
\caption{Extended mosaic plots for the Bundesliga and Hair-Eye-Color data (female students).}
\label{fig:both-mosaic-signif}
\end{center}
\end{figure}

\subsection{Implementation enhancements}

The current implementation of association and mosaic plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots (like mosaic matrices) as
they have been implemented using \textsf{R}'s base graphics engine where in
general plotting to relative coordinates is not supported.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some minor advantages and convenient
improvements. Second, the graphics parameters of the rectangles in
association and mosaic plots (like in almost all standard \textsf{R}
plots), such as color and line type, cannot be specified for each cell
by the user. To overcome this the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object being plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for the shading schemes
described in the previous section. Without this feature we would not be
able to specify the shading schemes described in the previous section
in such a flexible way.

\section{Multi-way tables}

Independence problems do not only occur in 2-way tables, although that
is an important special case, but they are also important in tables of
higher dimensionality and can follow much more complex patterns.
These are again defined based on the underlying table of theoretical cell probabilities
$(\pi_{ij\dots})$ with more than two dimensions. Models of interest include the
null hypotheses of:
\begin{itemize}
 \item total independence:
  $\pi_{ijk\dots} = \pi_{i++\dots}\pi_{+j+\dots}\pi_{++k\dots}\dots$
 \item conditional independence:
  $\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}$
 \item joint independence:
  $\pi_{ijk\dots} = \pi_{ij+\dots}\pi_{++k\dots}$
\end{itemize}
Classical non-graphical methods for these problems include the $\chi^2$
test, Fisher's exact test, the Cochran-Mantel-Haenzel test (for $2 \times 2 \times
K$-tables), and the analysis of log-linear models for more
complex settings.

As an example, two natural ways to use the visualization techniques described in the previous
sections would be to use (Trellis-like) conditioning plots or pairs plots
(like mosaic matrices) to visualize these more complex patterns of
independence.

Two ideas for the problem of conditional independence are briefly outlined
here and illustrated using the famous admissions data of the University of
California at Berkley (UCB) which is available in base \textsf{R}.
In this data, the question whether there is sex
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no sex discrimination
conditioned on the department---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
assocation plot in Figure \ref{fig:UCB-assoc-signif}.

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
tabplot(UCBAdmissions)
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

Similarly, the same data can be visualized using a mosaic matrix where
a conditional independence model is fitted in each plot (see Figure
\ref{fig:UCB-mosaic-signif}).

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
##grid.mosaicpairs(UCBAdmissions, type = "conditional", main = NULL)
grid.mosaicpairs(UCBAdmissions, type = "conditional", permute = FALSE)
@
\caption{Extended conditioned association plots for the UCB Admissions data.}
\label{fig:UCB-mosaic-signif}
\end{center}
\end{figure}

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables with some outlook to multi-way tables. The
extensions aim at improving the visualization by displaying both
the size of the residuals and the significance of a test for
independence and by using better color schemes. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

>>>>

\documentclass[10pt]{article}
\usepackage{dsc2003}
\usepackage{Sweave, graphicx}
\usepackage{Z}

@@

\SweaveOpts{engine=R}
\setkeys{Gin}{width=0.8\textwidth}

@@

\title{Visualizing Independence Using Extended Association Plots}
\author{\hfill David Meyer$^\dag$ \hfill Achim Zeileis$^\dag$ \hfill Kurt Hornik$^\ddag$ \hfill \hfill \\[0.2cm]
\dag {\it \small Institut f\"ur Statistik \& Wahrscheinlichkeitstheorie,
 Technische Universit\"at Wien}\\
\ddag {\it \small Institut f\"ur Statistik, Wirtschaftsuniversit\"at Wien}
}
\date{}
\maketitle

\begin{abstract}
Association plots---a visualization technique for the independence problem in 2-way
contingency tables---are extended in three directions:
\begin{enumerate}
 \item The visualization is enhanced by using colors for the
  importance of the residuals.
 \item Two extensions for the use of multi-way tables are proposed.
 \item The implementation in the \textsf{R} system is improved using a more modular
  design and allowing a more flexible specification of plotting parameters.
\end{enumerate}
\end{abstract}

@@

Given two categorical variables, or, equivalently, their contingency
table, one is often interested in investigating whether they are 
independent or not. Usually, the $\chi^2$ test statistic---the sum of the
squared Pearson residuals---is used as a measure of association.
Now, whenever the statistic is
significant, one might ask for a more detailed analysis on the basis
of the residuals themselves.
This task can greatly be sustained by the use of graphical
methods such as the association plot, for which we propose some
extensions: the use of colors for the shading of the residuals, two
possible extensions for multi-way tables, and a new, more flexible
implementation using {\sf R}'s alternative graphical system,
\texttt{grid}. The new functionality is provided by the {\sf R} package
\texttt{vcd}, inspired by the book `Visualizing Categorical Data'
\citep{vcd:Friendly:2000}.

\section{Association plots for 2-way tables} \label{sec:assocplots}

To assess independence of 2 categorical variables, usually
a 2-way contingency table is considered with cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
convenience, the number of observations is denoted $n = n_{++}$.

@@

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables. 
Now, when the $chi^2$ test statistic turns out to be significant for some data, 
it seems natural to go back to its components, i.e. the  residuals, for a more detailed analysis.

@@

Figure \ref{fig:assoc-classic1} shows the association plot
(produced by the implementation in {\sf R} `base') for the
variables `Hair' and `Eye' of the well-known `HairEyeColor' data
set (Hair and Eye colors of 264 male and 328 female statistics students). 
The biggest tiles are easy to depict: e.g., the tiles for 
brown and blue eyes, given blond hair, seem to represent important residuals. 
But for the others, their classification in important and less important 
categories seems not obvious. However, with some more information
visualized, this task could be facilitated.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-classic1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
base::assocplot(t(HairEyeColor[,,2]))
@
\caption{Association plot for `Hair' and `Eye' colors of female students.}
\label{fig:assoc-classic1}
\end{center}
\end{figure}

\section{Improved visualization through residual shading}

We propose to use a color-shading similar to Michael Friendly's
enhancements of the mosaic plot \cite[a graphical tool for the 
visualization of the observed frequencies in a contingency table; 
see][]{vcd:Hartigan+Kleiner:1984}:
this extension \citep{vcd:Friendly:1994} 
uses a color coding of the tiles to visualize deviations
(residuals) from a given log-linear model fitted to the table, that is, from
the expected frequencies under some independence hypothesis. 
Positive and negative signs of the residuals are coded by rectangles with
solid and dashed borders respectively. Furthermore, residuals can be
classified according to specified shading levels: per default,
residuals exceeding an absolute
value of 2 are shaded light blue and red respectively, those that even
exceed an absolute value of 4 are shaded with full saturation. The heuristic
behind this shading is that
the Pearson residuals are approximately standard normal which implies
that the highlighted cells are those with
residuals \emph{individually} significant at approximately the 5\% and 0.01\%
level. But the main purpose of the shading is not to visualize
significance but the \emph{pattern} of
deviation from independence \citep[p.~109]{vcd:Friendly:2000}.

Applying this color scheme to association plots similarly facilitates
the detection of (in)dependence patterns. Consider the `HairEyeColor' example
introduced in the previous section: Figure \ref{fig:assoc-improved1}
shows the same residuals than Figure \label{fig:assoc-classic1}, but
now, the color shading help us to sort the residuals into the 
three categories: important--less important--unimportant and thus
emphasizes the associational pattern in the underlying data.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved1,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(HairEyeColor[,,2], gp = gp.shading)
@
\caption{Extended association plot for `Hair' and `Eye' colors of female students.}
\label{fig:assoc-improved1}
\end{center}
\end{figure}

Another example is given in Figure \ref{fig:assoc-classic2}:
this time, we cross-tabulate the variables `Eye' and `Sex'.
The big tiles for brown Eyes suggest a strong asymmetry between
male and female students, and hence rejection of the independence
hypothesis. However, the $\chi^2$ test statistic is not significant at
a 5\% level: in this example, its value is 7.28, and the corresponding
$P$ value approximately $0.063$. Apparently, the relative size of the
tiles is insufficient to assess the ``strength'' of association
between two variables; again, information on the absolute size of the
residuals would be useful.
Using the extended shading (Figure \ref{fig:assoc-improved2}), we see at first glance
that \emph{no} residual is bigger than 2 (respectively smaller than -2 for
negative ones) because simply no tile is shaded. In fact, as we can depict from
the legend, the overall range of the residuals is approximately $[-1.5;1.5]$.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-classic2,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
base::assocplot(margin.table(HairEyeColor, 2:3))
@
\caption{Association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-classic2}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[htbp]
\begin{center}
<<assoc-improved,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
grid.assocplot(t(margin.table(HairEyeColor, 2:3)), gp = gp.shading)
@
\caption{Extended association plot for `Hair' and `Sex' in the `HairEyeColor' data.}
\label{fig:assoc-improved2}
\end{center}
\end{figure}

\section{Extension to multi-way tables}

Independence problems do not only occur in 2-way tables, but 
are also important in tables of higher dimensionality where they
can follow much more complex patterns.
These are again defined based on the underlying table of theoretical cell probabilities
$(\pi_{ij\dots})$ with more than two dimensions. Apart from the
analysis of all pairwise associations, models of interest include,
e.g., the null hypotheses of conditional independence:

\[\pi_{ijk\dots} = \pi_{i|k\dots}\pi_{j|k\dots}\]

A quick overview of the mutual independence structure of a multi-way table
can be given by a matrix of plots where each cell contains an
association plot of the corresponding row and column variables. Such an
\emph{association pairsplot} allows us to
quickly scan the association plots for all variable combinations. As
an example, we again use the HairEyeColor data (see Figure
\ref{fig:HEC-pairs}): only the cells corresponding to the variables
`Hair' and `Eye' contain an ``interesting'' association plot with
colored residuals.

\begin{figure}[htbp]
\begin{center}
<<UCB-mosaic-signif,fig=TRUE,echo=FALSE,height=8,width=8>>=
 panel.assocplot <- function (x, i, j, ..., axes, abbreviate) grid.assocplot(margin.table(x, c(i,j)), ..., panel = T, axis.labels = F, labels = F, main = NULL) 
grid.mosaicpairs(aperm(HairEyeColor), panel.upper= panel.assocplot, panel.lower = panel.assocplot, gp = gp.shading, main = NULL)
@
\caption{Extended association pairs for the `HairEyeColor' data.}
\label{fig:HEC-pairs}
\end{center}
\end{figure}

To visualize conditional independence, we could use conditioning
plots: like in trellis displays, we produce a separate plot
for each level of a conditioning variable (in the case of two
or more conditioning variables, for each level combination). 
This can be illustrated using the famous admissions data of the University of
California at Berkley (UCB) which is available in base \textsf{R}.
In this data, the question whether there is sex
discrimination at the UCB leads to the result that although women seem to
be disadvantaged at the aggregated level there is no sex discrimination
in the department strata---with the very exception of one department
in which women are {\em more} likely to be admitted than would be plausible
under independence. Exactly this is illustrated in the conditioning
association plot in Figure \ref{fig:UCB-assoc-signif}.

\begin{figure}[htbp]
\begin{center}
<<UCB-assoc-signif,fig=TRUE,echo=FALSE,height=10,width=10>>=
tabplot(UCBAdmissions, gp = gp.shading)
@
\caption{Extended conditioning association plots for the `UCB Admissions' data.}
\label{fig:UCB-assoc-signif}
\end{center}
\end{figure}

\section{Implementation enhancements}

The current implementation of association plots in \textsf{R} suffers
from two main disadvantages: First, it is not easy to recycle the
plots in conditioning plots or pairs plots as
they have been implemented using \textsf{R}'s base graphics engine where in
general plotting to relative coordinates is not supported.
The new implementation was written from scratch in \texttt{grid} offering
much more versatility amongst some minor advantages and convenient
improvements. Second, the graphics parameters of the rectangles in the
association plots (like in almost all standard \textsf{R}
plots), such as color and line type, cannot be specified for each cell
by the user. To overcome this, the current implementation in \texttt{vcd}
allows the user to specify either arrays of graphics parameters
of the same dimensionality as the object being plotted or a function
which computes these graphics parameters based on the original table
and its Pearson residuals. Functions are provided for shading schemes
like the one described in the previous section.

\section{Conclusion}
We suggest a set of enhancements for visualizing the independence
problem in 2-way tables using association plots. The
extensions aim at improving the visualization by displaying
the size of the residuals, and by extending the concept of association
plots to multi-way tables. Furthermore, a new
implementation is outlined based on the graphics package \texttt{grid}
which provides more modular design and more flexible specification
of graphical parameters.

Our work is still not at its end, though: for example, the plots for
multi-way tables currently do not use the same scale for all subplots,
and thus residuals of different subplots are not
comparable. Furthermore, the current color scheme only allows the
detection of ``patterns'' of independence. It would be nice, however, 
that the shading levels were
chosen such that a true visual test of independence can be performed,
that is, such that the presence of ``significant'' residuals lead to a
rejection of the hypothesis of independence. Current work includes the
development of such color schemes, based on an alternative test
statistic---the maximum of the absolute values of the pearson residuals.
